Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-08-18 docs/Manuscript/Manuscript.Rmd
update zotero keys + make tables consistent

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
Sys.sleep(10)
```

@@

We refer the interested reader to several previous reviews of CBR in medicine [@Gierl1998a; @Begum2011; @Choudhury2016], of measures of patient similarity [@Dai2020], and of uses of patient similarity in predictive models [@Welch2013; @Sharafoddini2017; @Parimbelli2018]. While the reviews of CBR focus on applications using health data, the patient similarity reviews encompass many additional types of data (various molecular -omics, genetic tests, medical images, laboratory tests, patient preferences, patient-reported outcomes, tracking devices, social media) and survey a much broader scope of models (exploratory analysis / dimension reduction / cluster analysis, risk evaluation / outcome prediction, clinical decision support / software tools). For the present review, we are interested in how similarity matching on patient-level health data is used to construct local cohorts for predictive modeling.

@@

```{r tab:synthesis, echo=FALSE, message=FALSE, warning=FALSE}
tab_synthesis <- readr::read_rds(here::here("tab/synthesis.rds"))
tab_synthesis |> 
  select(Citation, Task, Aim, Source, Type, Cases, Features) |> 
  # prevent numeric list columns from dominating the horizontal space
  mutate(across(
    c(-Citation, -Cases, -Features),
    ~ str_c(., "\\hspace{6em}")
  )) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:synthesis}",
      "Studies included in the method synthesis,",
      " arranged by the earliest the study is known to have been public."
    )
  )
```

```{r fig:year, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:year}
#| Number of publications in our sample each year."
knitr::include_graphics("../../fig/fig-years.png")
```

@@

```{r tab:composite, echo=FALSE, message=FALSE, warning=FALSE}
composite_studies <- read_rds(here::here("tab/composite.rds"))
composite_studies |> 
  transmute(
    # Citation = str_c("@", Citation),
    Citation = sapply(
      str_split(Citation, "; "),
      function(s) str_c("@", s, collapse = "; ")
    ),
    Elements = str_replace_all(Elements, ",", ";"),
    Terminology
  ) |> 
  mutate(across(
    everything(),
    # ~ str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) |>
  # mutate(across(c(Citation), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:composite}",
      "Methodological elements of studies included in the synthesis. "
    )
  )
```

@@

While all included studies were reported in scientific and medical journals, the vast majority were oriented toward clinical practice rather than medical research. For example, a 2006 study specifically evaluated the usefulness of CBR-based explanations for the purpose of decision support [@Doyle2006]. This study was part of a much larger literature on CBR systems and was included here despite relying on NN prediction because it used a non-traditional voting scheme to generate recommendations. A more recent study took essentially the same focus with respect to a proposed clinical risk prediction model, which amounted to CBR with a novel weighting scheme on predictors informed by expert consensus [@Fang2021a]. In both cases a prototype implementation was deployed in an experimental setting for evaluation.

@@

Toward detection and diagnosis, @Wyns2004 proposed a hybrid neural net–CBR system to distinguish (with confidence bounds) arthritic versus control patients, based on several histological features. @Nicolas2014a used collaborative multilabel CBR to subtype melanoma patients based on confocal and dermoscopy images. @Wang2019 used localized models built from a multi-type additive similarity measure to distinguish type 2 diabetic versus control populations. Along the way, @Song2006c, @Kasabov2010, and @Verma2015a took an iterative model-building approach to several tasks: predicting glomerular filtration rate, a key indicator of renal function, from demographic and physiological variables; identifying patients with colon cancer using a large number of gene expression measurements; and identifying patients with type 2 diabetes based on demographic, physiological, and genetic variables.

While several studies emphasized the potential or actual value to decision-making of their methods and tools, only one incorporated treatment decisions into their approach: By taking "decision points" as their units of analysis, @Tang2021d and @Ng2021b built not classifiers or predictors but comparative effectiveness models into a localized framework, providing for the first time in our sample explicitly prescriptive rather than descriptive clinical decision support.

@@

CBR can be understood as the opposite side of a trade-off with rule-based reasoning between model size and model complexity; @Doyle2006 describe their approach as "knowledge-light", in that "the cases do not contain explicit explanation structures; instead, explanation is achieved by comparison of the query case with retrieved cases". This means that the greatest performance and efficiency challenges in CBR have to do with the retrieval and revision phases in the R4 cycle. Several studies addressed these challenges: @Park2006, while excluded from the synthesis, were the earliest to propose that cohorts be bounded by a similarity threshold rather than by a number of cases, and this improved performance in their experiments. @CampilloGimenez2013 leveraged predictor weights obtained from logistic regression to inform the similarity calculation used in retrieval. @Lowsky2013 used non-parametric models to reduce the computational burden of revision (prediction). @Ma2020a proposed to improve efficiency along the entire R4 cycle, but in particular by only executing task-dependent steps in real time ("just-in-time learning"), and @Ng2021b and @Tang2021d partitioned multiple phases in the R4 cycle into offline and online components, only the latter of which would be performed in real time as new data are received.

Other studies addressed limitations of available tools. @Lopez2011 implemented a comprehensive CBR tool in response to the lack of general-purpose software, to enable coupling with other tools as well as to expedite development, experimentation, and uptake. Several other teams also set out to develop  more interoperable clinical support tools [@Elter2007; @Liang2015; @Ng2015; @Zhang2018a].

@@

The motivational and methodological unity of these studies does not reflect a unified research program. Besides the lack of any primary journal of record, we observed collaborations only among the authors of smaller contiguous programs, including applications of the TWNFI methodology [@Song2006c; @Verma2015a], individualized mortality prediction for ICU patients [@Lee2015; @Lee2017] and the use of more explanatory models to prioritize predictors or treatments for chronic disease [@Ng2015; @Tang2021d; @Ng2021b].

These programs used significantly varying terminology for common concepts, and in particular no common term is in use for what we here refer to as localized modeling; authors described this kind of method as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006c], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020a], and "precision cohort" analysis [@Wang2019; @Tang2021d; @Ng2021b]. We find uses of "targeted", "personalized", and "precision" generic and imprecise to this approach, while transductive inference is an established term for a broader set of methods in ML. Because local naturally contrasts with global, we propose "localized models" as a suitable term for this counterpart to globally-fitted models.

@@

The 25 studies included in our synthesis used composite techniques that we organized into two major types. One type, similarity learning, was used in 8 studies that tie in to the much larger literature on metric learning [@Yang2006]. These techniques are designed to detect structure in data, especially associations between predictors and responses, and to use this structure to inform the definition of a patient similarity measure. Seven studies used response values in training data sets to supervise similarity learning while two used unsupervised learning (Table \ref{tab:composite}). Based on the key properties of metric learning algorithms identified by @Bellet2014, most learned measures were non-linear while some were local, and some were optimized globally while others locally. However, not all qualified as metric learning, since the resulting measures would not necessarily satisfy the triangle inequality. Each then used its similarity measure to retrieve training cases relevant to each testing case. Once retrieved, most took a standard NN approach to generating predictions from the cohort; two exceptions [@Vilhena2016a; @Tang2021d] proceeded to fit predictive models to the retrieved cohorts and are discussed below.

@@

The similarity learning approaches took a variety of forms. Several studies used unsupervised similarity learning, for example the Mahalanobis distance [@Lowsky2013] and the Kolmogorov entropy-based distance [@Elter2007]. In particular, @Yearwood1997a defined similarity as a weighted sum of differences in predictor values and used linear regression to optimize the weights for the predictive accuracy of the cohort they retrieve. Others used supervised learning: @Song2006c proposed an iterative algorithm to optimize a set of fuzzy inference rules used to retrieve relevant cases and generate a prediction, using back-propagation on the rules’ parameters. Their method was used in several later studies returned by our search, which are not discussed here because they did not originate the technique. @Nicolas2014a explicitly appeal to a supervised distance metric learning technique [@Xing2002] to obtain a similarity measure on their set of dermoscopy and confocal images that most effectively separates malignant from benign melanoma tumors. We will discuss the remaining uses of supervised similarity learning shortly.

@@

In addition to @Lowsky2013, @Liang2015, @Ng2015, @Lee2017, @Vilhena2016a, @Tang2021d, and @Ng2021b, five other studies employed localized models: @Mariuzzi1997, @Lee2015, @Verma2015a, @Wang2019, and @Ma2020a. In most cases these models were purely predictive in application; of the exceptions, @Mariuzzi1997 produced purely descriptive models of survival outcomes, which were evaluated for their precision rather than their accuracy, while @Ng2015 used generalized regression models descriptively as well as predictively, as did some of the same authors [@Tang2021d; @Ng2021b] later. The most common approach to cohort selection was to optimize a size threshold via manual exploration [@Mariuzzi1997; @Lee2015; @Ng2015; @Lee2017; @Wang2019; @Vilhena2016a] or via cross-validation [@Lowsky2013; @Verma2015a]. Further in the former direction, one study [@Ma2020a] set a fixed cohort size while two [@Liang2015; @Tang2021d; @Ng2021b] devised more sophisticated algorithms to balance multiple desiderata including predictive accuracy.

Several themes emerged from this sample. Foremost was the optimization of retrieved cohort sizes for some measure of model performance. Despite the early recommendation by @Park2006 to threshold cohorts by similarity rather than by cardinality, only the one previously published study @Mariuzzi1997 and the most recent study [@Tang2021d; @Ng2021b] in this corpus took this approach. Viewed as a hyperparameter of the individualized modeling approach, cohort size can be treated in the same way as neighborhood size in NN prediction (viewed here as a special case of localized modeling wherein the model is a simple summary statistic), so we will not discuss it further.

An alternative approach was to optimize multiple hyperparameters together: @Kasabov2010 proposed an iterative algorithm to settle on an optimal number and set of predictors as well as number of training cases (comprising the retrieved cohort) that was later used by @Liang2015. More recently, @Tang2021d and @Ng2021b proposed a three-step process for cohort selection that filtered by exact match for one subset of predictors, used domain-informed similarity measures to rank these, and optimized the similarity threshold for a trade-off between cohort size and a measure of bias called cohort balance. Notably, though their similarity measure was supervised, this optimization process was not.

With the exception of the descriptive survival models of @Mariuzzi1997, only one thread in this corpus concerned itself with the interpretability of localized models: @Ng2015 fit logistic regression models to similarity-based cohorts and examined not only their predictive performance but the localized sets of largest and most detectable predictors, termed risk profiles, and how they differed across the testing population. This approach informed their later use of localized comparative effectiveness–style models to recommend courses of treatment at decision points during a monitored patent’s stay [@Tang2021d; @Ng2021b]. These studies suggest much wider potential for localized real-world evidence generation, which has long been promoted as a promise of advances in clinical and health informatics [@Longhurst2014].

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
Sys.sleep(10)
tab_synthesis <- readr::read_rds(here::here("data/synthesis.rds"))
composite_studies <- readr::read_rds(here::here("data/composite.rds"))
citation_id <- readr::read_rds(here::here("data/citations.rds"))
```

@@

We refer the interested reader to several previous reviews of CBR in medicine [@Gierl1998; @Begum2011; @Choudhury2016], of measures of patient similarity [@Dai2020], and of uses of patient similarity in predictive models [@Welch2013; @Sharafoddini2017; @Parimbelli2018]. While the reviews of CBR focus on applications using health data, the patient similarity reviews encompass many additional types of data (various molecular -omics, genetic tests, medical images, laboratory tests, patient preferences, patient-reported outcomes, tracking devices, social media) and survey a much broader scope of models (exploratory analysis / dimension reduction / cluster analysis, risk evaluation / outcome prediction, clinical decision support / software tools). For the present review, we are interested in how similarity matching on patient-level health data is used to construct local cohorts for predictive modeling.

@@

```{r tab:synthesis, echo=FALSE, message=FALSE, warning=FALSE}
tab_synthesis |> 
  # combine citations of the same study
  left_join(citation_id, by = "Citation") |> 
  group_by(across(c(-Number, -Citation, -Date))) |> 
  summarize(Date = min(Date), Citation = str_c(Citation, collapse = "; ")) |> 
  ungroup() |> 
  arrange(Date) |> 
  select(Citation, Task, Aim, Source, Type, Cases, Features) |> 
  # format citations for Markdown-LaTeX
  # mutate(Citation = str_c("[@", Citation, "]")) |> 
  mutate(Citation = sapply(
    str_split(Citation, "; "),
    function(s) str_c("@", s, collapse = "; ")
  )) |> 
  # prevent numeric list columns from dominating the horizontal space
  mutate(across(
    c(-Citation, -Cases, -Features),
    ~ str_c(., "\\hspace{6em}")
  )) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:synthesis}",
      "Studies included in the method synthesis,",
      " arranged by the earliest the study is known to have been public."
    )
  )
```

```{r fig:year, fig.align='center', out.width="75%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:year}
#| Number of publications in our sample each year."
knitr::include_graphics("../../fig/fig-years.png")
```

@@

```{r tab:composite, echo=FALSE, message=FALSE, warning=FALSE}
composite_studies |> 
  transmute(
    # Citation = str_c("@", Citation),
    Citation = sapply(
      str_split(Citation, "; "),
      function(s) str_c("@", s, collapse = "; ")
    ),
    Elements = str_replace_all(Elements, ",", ";"),
    Terminology
  ) |> 
  mutate(across(
    everything(),
    # ~ str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) |>
  # mutate(across(c(Citation), ~ str_c(., "\\hspace{6em}"))) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:composite}",
      "Methodological elements of studies included in the synthesis. "
    )
  )
```

@@

While all included studies were reported in scientific and medical journals, the vast majority were oriented toward clinical practice rather than medical research. For example, a 2006 study specifically evaluated the usefulness of CBR-based explanations for the purpose of decision support [@Doyle2006a]. This study was part of a much larger literature on CBR systems and was included here despite relying on NN prediction because it used a non-traditional voting scheme to generate recommendations. A more recent study took essentially the same focus with respect to a proposed clinical risk prediction model, which amounted to CBR with a novel weighting scheme on predictors informed by expert consensus [@Fang2021a]. In both cases a prototype implementation was deployed in an experimental setting for evaluation.

@@

Toward detection and diagnosis, @Wyns2004a proposed a hybrid neural net–CBR system to distinguish (with confidence bounds) arthritic versus control patients, based on several histological features. @Nicolas2014a used collaborative multilabel CBR to subtype melanoma patients based on confocal and dermoscopy images. @Wang2019 used localized models built from a multi-type additive similarity measure to distinguish type 2 diabetic versus control populations. Along the way, @Song2006c, @Kasabov2010, and @Verma2015a took an iterative model-building approach to several tasks: predicting glomerular filtration rate, a key indicator of renal function, from demographic and physiological variables; identifying patients with colon cancer using a large number of gene expression measurements; and identifying patients with type 2 diabetes based on demographic, physiological, and genetic variables.

While several studies emphasized the potential or actual value to decision-making of their methods and tools, only one incorporated treatment decisions into their approach: By taking "decision points" as their units of analysis, @Tang2021c and @Ng2021b built not classifiers or predictors but comparative effectiveness models into a localized framework, providing for the first time in our sample explicitly prescriptive rather than descriptive clinical decision support.

@@

CBR can be understood as the opposite side of a trade-off with rule-based reasoning between model size and model complexity; @Doyle2006a describe their approach as "knowledge-light", in that "the cases do not contain explicit explanation structures; instead, explanation is achieved by comparison of the query case with retrieved cases". This means that the greatest performance and efficiency challenges in CBR have to do with the retrieval and revision phases in the R4 cycle. Several studies addressed these challenges: @Park2006, while excluded from the synthesis, were the earliest to propose that cohorts be bounded by a similarity threshold rather than by a number of cases, and this improved performance in their experiments. @CampilloGimenez2013 leveraged predictor weights obtained from logistic regression to inform the similarity calculation used in retrieval. @Lowsky2013 used non-parametric models to reduce the computational burden of revision (prediction). @Ma2020a proposed to improve efficiency along the entire R4 cycle, but in particular by only executing task-dependent steps in real time ("just-in-time learning"), and @Ng2021b and @Tang2021c partitioned multiple phases in the R4 cycle into offline and online components, only the latter of which would be performed in real time as new data are received.

Other studies addressed limitations of available tools. @Lopez2011a implemented a comprehensive CBR tool in response to the lack of general-purpose software, to enable coupling with other tools as well as to expedite development, experimentation, and uptake. Several other teams also set out to develop  more interoperable clinical support tools [@Elter2007a; @Liang2015a; @Ng2015; @Zhang2018a].

@@

The motivational and methodological unity of these studies does not reflect a unified research program. Besides the lack of any primary journal of record, we observed collaborations only among the authors of smaller contiguous programs, including applications of the TWNFI methodology [@Song2006c; @Verma2015a], individualized mortality prediction for ICU patients [@Lee2015; @Lee2017] and the use of more explanatory models to prioritize predictors or treatments for chronic disease [@Ng2015; @Tang2021c; @Ng2021b].

These programs used significantly varying terminology for common concepts, and in particular no common term is in use for what we here refer to as localized modeling; authors described this kind of method as "targeted prognosis" [@Mariuzzi1997], "transductive inference" [@Song2006c], "personalized decision support" [@Lee2015], "personalized (predictive) models" [@Liang2015a, in contrast to "local models"; @Ng2015; @Wang2019; @Ma2020a], and "precision cohort" analysis [@Wang2019; @Tang2021c; @Ng2021b]. We find uses of "targeted", "personalized", and "precision" generic and imprecise to this approach, while transductive inference is an established term for a broader set of methods in ML. Because local naturally contrasts with global, we propose "localized models" as a suitable term for this counterpart to globally-fitted models.

@@

The 25 studies included in our synthesis used composite techniques that we organized into two major types. One type, similarity learning, was used in 8 studies that tie in to the much larger literature on metric learning [@Yang2006]. These techniques are designed to detect structure in data, especially associations between predictors and responses, and to use this structure to inform the definition of a patient similarity measure. Seven studies used response values in training data sets to supervise similarity learning while two used unsupervised learning (Table \ref{tab:composite}). Based on the key properties of metric learning algorithms identified by @Bellet2014, most learned measures were non-linear while some were local, and some were optimized globally while others locally. However, not all qualified as metric learning, since the resulting measures would not necessarily satisfy the triangle inequality. Each then used its similarity measure to retrieve training cases relevant to each testing case. Once retrieved, most took a standard NN approach to generating predictions from the cohort; two exceptions [@Vilhena2016a; @Tang2021c] proceeded to fit predictive models to the retrieved cohorts and are discussed below.

@@

The similarity learning approaches took a variety of forms. Several studies used unsupervised similarity learning, for example the Mahalanobis distance [@Lowsky2013] and the Kolmogorov entropy-based distance [@Elter2007a]. In particular, @Yearwood1997a defined similarity as a weighted sum of differences in predictor values and used linear regression to optimize the weights for the predictive accuracy of the cohort they retrieve. Others used supervised learning: @Song2006c proposed an iterative algorithm to optimize a set of fuzzy inference rules used to retrieve relevant cases and generate a prediction, using back-propagation on the rules’ parameters. Their method was used in several later studies returned by our search, which are not discussed here because they did not originate the technique. @Nicolas2014a explicitly appeal to a supervised distance metric learning technique [@Xing2002] to obtain a similarity measure on their set of dermoscopy and confocal images that most effectively separates malignant from benign melanoma tumors. We will discuss the remaining uses of supervised similarity learning shortly.

@@

In addition to @Lowsky2013, @Liang2015a, @Ng2015, @Lee2017, @Vilhena2016a, @Tang2021c, and @Ng2021b, five other studies employed localized models: @Mariuzzi1997, @Lee2015, @Verma2015a, @Wang2019, and @Ma2020a. In most cases these models were purely predictive in application; of the exceptions, @Mariuzzi1997 produced purely descriptive models of survival outcomes, which were evaluated for their precision rather than their accuracy, while @Ng2015 used generalized regression models descriptively as well as predictively, as did some of the same authors [@Tang2021c; @Ng2021b] later. The most common approach to cohort selection was to optimize a size threshold via manual exploration [@Mariuzzi1997; @Lee2015; @Ng2015; @Lee2017; @Wang2019; @Vilhena2016a] or via cross-validation [@Lowsky2013; @Verma2015a]. Further in the former direction, one study [@Ma2020a] set a fixed cohort size while two [@Liang2015a; @Tang2021c; @Ng2021b] devised more sophisticated algorithms to balance multiple desiderata including predictive accuracy.

Several themes emerged from this sample. Foremost was the optimization of retrieved cohort sizes for some measure of model performance. Despite the early recommendation by @Park2006 to threshold cohorts by similarity rather than by cardinality, only the one previously published study @Mariuzzi1997 and the most recent study [@Tang2021c; @Ng2021b] in this corpus took this approach. Viewed as a hyperparameter of the individualized modeling approach, cohort size can be treated in the same way as neighborhood size in NN prediction (viewed here as a special case of localized modeling wherein the model is a simple summary statistic), so we will not discuss it further.

An alternative approach was to optimize multiple hyperparameters together: @Kasabov2010 proposed an iterative algorithm to settle on an optimal number and set of predictors as well as number of training cases (comprising the retrieved cohort) that was later used by @Liang2015a. More recently, @Tang2021c and @Ng2021b proposed a three-step process for cohort selection that filtered by exact match for one subset of predictors, used domain-informed similarity measures to rank these, and optimized the similarity threshold for a trade-off between cohort size and a measure of bias called cohort balance. Notably, though their similarity measure was supervised, this optimization process was not.

With the exception of the descriptive survival models of @Mariuzzi1997, only one thread in this corpus concerned itself with the interpretability of localized models: @Ng2015 fit logistic regression models to similarity-based cohorts and examined not only their predictive performance but the localized sets of largest and most detectable predictors, termed risk profiles, and how they differed across the testing population. This approach informed their later use of localized comparative effectiveness–style models to recommend courses of treatment at decision points during a monitored patent’s stay [@Tang2021c; @Ng2021b]. These studies suggest much wider potential for localized real-world evidence generation, which has long been promoted as a promise of advances in clinical and health informatics [@Longhurst2014].
